Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-08-01 sports_analytics.Rmd
More Minor edits.

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

@@

XX Seems massive

@@

The concept of expected points in American football has many applications.
For example, one of the most notable and well-studied topic is the evaluation of 4th down strategy.
There is near universal consensus in the literature that NFL teams have been too conservative when making 4th down decisions.

@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making). 
He concludes that teams don't go for it enough if they are trying to maximize their probability of winning the game.

Numerous papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy.
In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.
In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

Recently, @lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.
Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.
They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.

@@

In addition, @Alamar2010 implements an expected points framework to examine play calling in the NFL. 
However, rather than assessing each play on its own, the author evaluates the play in the context of the drive. 
Based on play-by-play data from 2005 through 2008, he determines that teams are under-utilizing passing plays in some situations.

@@

Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated with the help of tracking data. 
Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. 
This high-resolution data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game. 
This is a great leap forward from older resources such as traditional box-score results and play-by-play data.

>>>>

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

@@

<!-- XX Seems massive -->

@@

The concept of expected points in American football has many applications.  One of the most notable and well-studied topic is the evaluation of 4th down strategy.
There is near universal consensus in the literature that NFL teams have been too conservative in the past when making 4th down decisions.

@Romer2006 examines 4th down decisions in the NFL using expected points by focusing only on examples from the first quarter of a game (to avoid issues with end-of-half and end-of-game decision making). 
They concluded that teams don't go for it enough if teams are trying to maximize their probability of winning the game.

Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.  In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@lopez2020bigger presents an introduction to NFL tracking data, and as an example, examines 4th down behavior.  Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go. They demonstrate that teams are still conservative when choosing to go for it or punting on 4th down, but the effect may be overstated.

@@

@Alamar2010 implements an expected points framework to examine play calling in the NFL. However, rather than assessing each play on its own, they evaluate the play in the context of the drive. 
Based on play-by-play data from 2005 through 2008, they determine that teams are under-utilizing passing plays in some situations.

@@

Even in sports where the concept of a state is more difficult to define, the value of a possession can be estimated with the help of tracking data.  Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. This high-resolution data allows researchers to produce advanced analyses of the captured spatiotemporal information and better understand the game.  This is a great leap forward from older resources such as traditional box-score results and play-by-play data.
